---
title: "Milestone Report Coursera Data Science Capstone"
author: "Jose Antonio (joseantonio@me.com"
date: "10/11/2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1-INTRODUCTION

This Milestone Report is about exploratory data analysis of the Capstone Project of the Data Science Coursera specialization.
Coursera and SwitfKey are partnering at this project, that apply data science in area of natural language.
The project use a large text corpus of documents to predict the next word on preceding input.  
The data will be extracted and cleaned from these documents to be used with Shiny application.
Here, we have some informations about the corpus of data and prepare a plan to create the predictive model.

## 2-R LIBRARIES

Some R libraries used at this project.

```{r, echo=TRUE}
library(stringi) # stats files
library(NLP); library(openNLP)
library(tm) # Text mining
library(rJava)
library(RWeka) # to create unigrams, bigrams, trigrams
library(RWekajars)
library(SnowballC) # Stemming
library(qdap); library(qdapTools); library(qdapRegex); library(qdapDictionaries)
library(ggplot2) #visualization
library(RColorBrewer) # Color palettes
library(wordcloud) # word cloud generator (not used)
library(ngramrr)
library(magrittr)

## in study
library(quanteda) # faster cleaning/processing than RWeka or tm packages, can also creat n-grams in the same call.  

```

## 3 - LOOKING FOR THE DATA 
The data is from HC Corpora with access to 4 languages, but only English will be used. The dataset has 3 files.

 * en_US.blogs.txt
 * en_US.news.txt
 * en_US.twitter.txt.

The data was loaded from Coursera Link to local machine and will be read from local disk. 

```{r, echo=TRUE, message=FALSE, warning=FALSE}
blogsURL <- file("~/Desktop/DATA-SCIENCE-CAPSTONE/en_US.blogs.txt", open="rb") # open for reading in binary mode
blogs <- readLines(blogsURL, encoding = "UTF-8", skipNul=TRUE)

newsURL <- file("~/Desktop/DATA-SCIENCE-CAPSTONE/en_US.news.txt", open = "rb") # open for reading in binary mode
news <- readLines(newsURL, encoding = "UTF-8", skipNul=TRUE)

twitterURL <- file("~/Desktop/DATA-SCIENCE-CAPSTONE/en_US.twitter.txt", open = "rb") # open for reading in binary mode
twitter <- readLines(twitterURL, encoding = "UTF-8", skipNul=TRUE)

rm(blogsURL, newsURL, twitterURL)
```

## 4 - SOME DATA STATISTICS

Lets evaluate the Data loaded with

```{r, echo=TRUE}
## Size of Files
file.info("~/Desktop/DATA-SCIENCE-CAPSTONE/en_US.blogs.txt")$size / 1024^2 # Megabytes
file.info("~/Desktop/DATA-SCIENCE-CAPSTONE/en_US.news.txt")$size  / 1024^2 # Megabytes
file.info("~/Desktop/DATA-SCIENCE-CAPSTONE/en_US.twitter.txt")$size / 1024^2 # Megabytes

## Number of lines
length(blogs) # 899,288 lines
length(news)  # 1,010,242 lines
length(twitter) # 2,360,148

## Counting the Words
sum(stri_count_words(blogs)) # words at blogs = 37,546,246
sum(stri_count_words(news))  # words at news =  34,762,395
sum(stri_count_words(twitter)) # words at twitter = 30,093,410

## The length of the longest line seen in any of the three en_US data sets: (question 3 of Quiz 1)
max(nchar(blogs)) # [1] 40,833
max(nchar(news))  # [1] 11,384 
max(nchar(twitter)) # [1] 140
```


### 4.1 - DATA SUMMARY OBSERVED

* Each file has more than 150 MB.
* The number of words is more than 30 million per file.
* Twitter is the big file with more lines, and less words per line (as expected 140 lines limited).
* Blogs are text file with sentences and has the longest line with 40,833 characters.
* News are text file with senteces.
* Twitter are text file with sentences until 140 characters as would expected and with 2,360,148 number of lines.


## 5 - DATA SAMPLE (subset)

To make the analyse and prediction, the data is very large, and could have poor perfomance in mobile.
So, we must create a subset of the data considering the limited resources for test and application.
A data sample was generated from the three files observed.


```{r, eval=FALSE}
sTwitter <- sample(twitter, size = 5000, replace = TRUE)
sBlogs <- sample(blogs, size = 5000, replace = TRUE)
sNews <- sample(news, size = 5000, replace = TRUE)
sampleTotal <- c(sTwitter, sBlogs, sNews)
length(sampleTotal)
rm(sBlogs, sNews, sTwitter)
rm(blogs, news, twitter)
```

The Sample chooses has 15,000 lines, 5,000 from each of the files.

```{r, eval=FALSE, echo=FALSE}
## Saving the Sample Total
writeLines(sampleTotal, "~/Desktop/DATA-SCIENCE-CAPSTONE/sampleTotal.txt")
rm(sampleTotal)
```


## 6 - CORPUS AND CLEANING THE DATA

The final text data needs to be cleaned to be used at the word prediction.
The objective is to create a cleaned Corpus file or sample of text. 
This Corpus will be cleaned using methods as removing whitespaces, removing numbers, UTR, punctuation and so on.
The library used here is TM that loads the corpus into memory and allow calls to the methods to clean the data.

Profanity words have to be erased from the dataset.

### 6.1 - Clean the Data
```{r, eval=FALSE, echo=TRUE}
## Using the TM Package to clean the Corpus Text
textCon <- file("~/Desktop/DATA-SCIENCE-CAPSTONE/sampleTotal.txt")
textCorpus <- readLines(textCon)
## Tm readin the text as lists.
textCorpus <- Corpus(VectorSource(textCorpus))
rm(textCon)

## Using the TM Package to clean the text
textCorpus <- tm_map(textCorpus, content_transformer(function(x) iconv(x, to="UTF-8", sub="byte")),mc.cores=2)

#1 removing number
textCorpus <- tm_map(textCorpus, content_transformer(removeNumbers))

#2 removing punctuation
textCorpus <- tm_map(textCorpus, content_transformer(removePunctuation), preserve_intra_word_dashes=TRUE)

#3 converting to lowercase
textCorpus <- tm_map(textCorpus, content_transformer(tolower), lazy = TRUE)

#4 removing stop words in English (a, as, at, so, etc.)
textCorpus <- tm_map(textCorpus, removeWords, stopwords("english"))

#5 Removing common word endings (e.g., ???ing???, ???es???, ???s???)
textCorpus <- tm_map(textCorpus, stemDocument) 

#6 Stripping unnecesary whitespace from document
textCorpus <- tm_map(textCorpus, stripWhitespace) 

# removing URLs 
removeURL <- function(x) gsub("http[[:alnum:]]*", "", x) 
textCorpus <- tm_map(textCorpus, content_transformer(removeURL))
rm(removeURL)

# Removing Profanity Words
# profanityWords <- read.table("~/Desktop/Capstone2/MilestoneReport/profanityfilter.txt", header = FALSE)
# cleanSample <- tm_map(cleanSample, removeWords, profanityWords)
      
## Convert all to plain text document
textCorpus <- tm_map(textCorpus, PlainTextDocument) 

```

### 6.1 - Saving the final Corpus as .RDS

```{r, eval=FALSE, echo=FALSE}
## Saving the final corpus
saveRDS(textCorpus, file = "~/Desktop/DATA-SCIENCE-CAPSTONE/finalCorpus.RDS")
finalCorpus <- readRDS("~/Desktop/DATA-SCIENCE-CAPSTONE/finalCorpus.RDS")
finalCorpusDF <-data.frame(text=unlist(sapply(finalCorpus,`[`, "content")),stringsAsFactors = FALSE)
```


## 7 - TOKENIZATION

Tokenize the corpus to determine the words and phrases frequency.
In Natural Language Processing (NLP) an *n*-gram is a contiguous sequence of n items from a given sequence of text or speech.
Unigrams are single words. Bigrams are two words combinations. Trigrams are three word combinations.
The following function is used to extract 1-grams, 2-grams and 2-grams from the cleaned text corpus.
Generating uni, bi and tri N-gram.

By the usage of the tokenizer function for the *n*-grams a distribution of the following top 10 words and word combinations can be inspected. 
Unigrams are single words, while bigrams are two word combinations and trigrams are three word combinations.



## 8 - NEW STEPS






